---
title: "El Cerrito Police Incident Analysis 2019-2025"
author: "Ira Sharenow"
date: "August 11, 2025"
output: 
  word_document:
    number_sections: false
    fig_caption: true
---

```{r setup, include=FALSE}
# Global settings
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)
```


```{r libraries}
# Use the same library set as the fire project
library(readxl)
library(dplyr)
library(lubridate)
library(ggplot2)
library(stringr)
library(forcats)
library(tidyr)
library(flextable)
library(scales)
library(writexl)
library(tidygeocoder)
library(geosphere)
library(ggmap)
library(viridis)
```


```{r global-theme}
# Consistent minimalist look across all charts (including heat maps)
theme_set(
  theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
    )
)
```

```{r global_flextable function}
# Global Flextable Function (from fire project) — with minor safety tweaks
# Purpose: Uniform flextable styling with safe guards for optional columns
# Notes:
# - Does NOT assume a YEAR column exists; skips if absent.  # added
# - Only formats columns that are present in `tbl`.          # added

table_fire_style <- function(tbl,
                             label_map,
                             currency_cols = NULL,
                             percent_cols = NULL,
                             color_col = NULL,
                             color_domain = NULL,
                             caption = NULL,
                             footer = NULL) {

  ft <- flextable::flextable(tbl) |>
    flextable::set_header_labels(values = label_map)

  # ---- Safe year formatting (only if a YEAR/year column exists) ----  # added
  possible_years <- c("YEAR", "year")
  year_col <- possible_years[possible_years %in% names(tbl)]
  if (length(year_col) == 1) {
    yc <- year_col[1]
    if (is.numeric(tbl[[yc]])) {
      ft <- ft |> flextable::colformat_num(j = yc, digits = 0, big.mark = "")
    } else {
      ft <- ft |> flextable::colformat_char(j = yc)
    }
  }
  # -------------------------------------------------------------------

  # Currency columns (only those that exist)
  if (!is.null(currency_cols)) {
    cc <- intersect(currency_cols, names(tbl))
    if (length(cc)) {
      ft <- ft |> flextable::colformat_num(j = cc, digits = 0, big.mark = ",", prefix = "$")
    }
  }

  # Percent columns (only those that exist)
  if (!is.null(percent_cols)) {
    pc <- intersect(percent_cols, names(tbl))
    if (length(pc)) {
      ft <- ft |> flextable::colformat_num(j = pc, digits = 1, suffix = "%")
    }
  }

  # Optional color emphasis on a numeric column
  if (!is.null(color_col) && !is.null(color_domain) && color_col %in% names(tbl)) {
    color_values <- tbl[[color_col]]
    color_fn <- scales::col_numeric(palette = c("#FFCCCC", "#990000"), domain = color_domain)
    color_vec <- ifelse(!is.na(color_values) & color_values > 0, color_fn(color_values), NA)
    ft <- ft |> flextable::color(j = color_col, color = color_vec)
  }

  ft |>
    flextable::autofit() |>
    flextable::theme_booktabs() |>
    flextable::align(align = "right", part = "body") |>
    flextable::align(align = "center", part = "header") |>
    flextable::bold(part = "header") |>
    flextable::fontsize(size = 11, part = "all") |>
    flextable::set_caption(caption) |>
    flextable::add_footer_lines(footer)
}


```

```{r global-functions}
# Global Charting Function (from fire project)
# Bar chart (simple, readable labels)
create_bar_chart <- function(data, x_var, y_var, title, x_label, y_label) {
  ggplot(data, aes_string(x = x_var, y = y_var)) +
    geom_col(fill = "steelblue") +
    geom_text(aes_string(label = y_var), vjust = -0.3, size = 3.5) +
    labs(title = title, x = x_label, y = y_label) +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
}

# Heat map (assumes data already has longitude/latitude)
create_heat_map <- function(data, lon = "longitude", lat = "latitude",
                            title = "Incident Density Heat Map",
                            subtitle = NULL,
                            bins = 40, alpha = 0.8) {
  if (!all(c(lon, lat) %in% names(data))) {
    return(ggplot() + annotate("text", x = 0, y = 0,
                               label = sprintf("Missing columns: %s / %s", lon, lat),
                               size = 4) + theme_void())
  }
  d <- data[!is.na(data[[lon]]) & !is.na(data[[lat]]), , drop = FALSE]
  if (!nrow(d)) {
    return(ggplot() + annotate("text", x = 0, y = 0,
                               label = "No points with coordinates.", size = 4) +
             theme_void())
  }

  ggplot() +
    stat_density_2d(
      data = d,
      aes(x = .data[[lon]], y = .data[[lat]], fill = after_stat(level)),
      geom = "polygon", contour = TRUE, alpha = alpha, bins = bins
    ) +
    scale_fill_viridis_c(name = "Incident Density") +
    labs(title = title, subtitle = subtitle) +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
}
```


```{r geocode_start}

# Point directly to the file you produced after geocoding yesterday
data_dir   <- "D:/Documents/Employment/2025 job search/Project 2025/el-cerrito-police-report/data/combined"
active_csv <- file.path(data_dir, "el-cerrito-police-with-coords.csv")  # or "el-cerrito-police-usable-coords.csv"

if (!file.exists(active_csv)) {
  stop("Geocoded CSV not found at: ", active_csv)
}

df <- readr::read_csv(active_csv, show_col_types = FALSE)

# If the file still has original headers, rename once; otherwise this is a no-op
if ("Event Number" %in% names(df)) {
  df <- dplyr::rename(
    df,
    event_number            = `Event Number`,
    call_for_service        = `Call for Service`,
    received_date           = `Received Date`,
    first_unit_arrived_time = `First Unit Arrived Time`,
    cleared_date            = `Cleared Date`,
    address                 = `Address`
  )
}

# Light clean (idempotent)
df <- df |>
  dplyr::distinct() |>
  dplyr::filter(!is.na(event_number)) |>
  dplyr::mutate(
    call_for_service = stringr::str_trim(stringr::str_to_lower(call_for_service)),
    address          = stringr::str_squish(address)
  )

# Parse dates if they are not already POSIXct
if (!inherits(df$received_date, "POSIXt")) {
  df <- df |>
    dplyr::mutate(
      received_date = lubridate::parse_date_time(received_date, orders = c("mdy HM", "mdy HMS", "Y-m-d H:M:S")),
      first_unit_arrived_time = lubridate::parse_date_time(first_unit_arrived_time, orders = c("mdy HM", "mdy HMS", "Y-m-d H:M:S")),
      cleared_date = lubridate::parse_date_time(cleared_date, orders = c("mdy HM", "mdy HMS", "Y-m-d H:M:S"))
    )
}

```

\newpage  
## Introduction

First I would like to thank the El Cerrito Police Department for providing police incident data from January 1, 2019-June 30, 2025. But be sure to read the technical note to see how I cleaned the data.

I have two goals for this project. One goal is to provide standard summary statistics such as number of incidents, seasonality, and so forth. The other goal is to focus in on specific landmarks. In particular, El Cerrito residents are debating whether to spend millions of dollars per year to move the library to an area near El Cerrito Plaza. Many points have been made on both sides of the issue. This report looks at police incident data, which no one else has looked at so far.

In many cases, there will be a chart and a table accompanied by an explanation.

**Main takeaways**  
-- Incident counts vary considerably by location with a large number near San Pablo Avenue.  
-- The proposed new library location is a significant incident hotspot.  
-- Incident counts declined considerably during the Covid years of 2020-2022.  
-- There is not much seasonal variability.  


\newpage




```{r read_data}
# Purpose: Read cleaned incidents, then filter by updated call_for_service list.
# Keeps the SAME object name: df_filtered_clean_updated

# 1) Read your cleaned/geocoded dataset (already saved previously)
df_filtered_clean_updated <- readRDS(
  "D:/Documents/Employment/2025 job search/Project 2025/el-cerrito-police-report/data/combined/df_filtered_clean_updated.rds"
)

# 2) Read the updated list of incident types
types_path <- "D:/Documents/Employment/2025 job search/Project 2025/el-cerrito-police-report/data/combined/types of incidents filtered.csv"
incident_types_raw <- readr::read_csv(types_path, show_col_types = FALSE)

# Use the first column as the list (keeps it simple). Make it lower/trim to match your main data cleaning.
incident_types <- incident_types_raw[[1]] |>
  as.character() |>
  stringr::str_trim() |>
  stringr::str_to_lower() |>
  unique()

# 3) Normalize call_for_service in the main data and filter IN PLACE
df_filtered_clean_updated <- df_filtered_clean_updated |>
  dplyr::mutate(call_for_service = stringr::str_to_lower(stringr::str_trim(call_for_service))) |>
  dplyr::filter(!is.na(call_for_service), call_for_service %in% incident_types)
```

**Heatmap of Police Incidents Hotspots**

The first chart is a heatmap which shows the relative frequencies of incidents in the different parts of El Cerrito. In addition, I chose a number of landmarks which I thought would be of interest to El Cerrito residents. Interestingly, the chart and the first table show that the proposed location for a new library had significantly more incidents than any of the other landmarks.  
Given the significance of the proposed library site, I further explored and listed the counts of the types of incidents near the proposed library site and compared to the current site.    

The counts are within 500 feet of each landmark.  


```{r chart1_heatmap_total_incidents}
# Purpose: First heatmap (2017–2025) using the filtered df_filtered_clean_updated and Excel landmarks.

# Ensure received_date is POSIXct if we’re filtering by year
if (!inherits(df_filtered_clean_updated$received_date, "POSIXt")) {
  df_filtered_clean_updated <- df_filtered_clean_updated %>%
    dplyr::mutate(received_date = lubridate::parse_date_time(
      received_date, orders = c("mdy HM","mdy HMS","Y-m-d H:M:S")
    ))
}

# --- Landmarks: read Excel and rename columns in a straightforward way ---
lm_path <- "D:/Documents/Employment/2025 job search/Project 2025/el-cerrito-police-report/data/combined/landmark lat and long.xlsx"
lm_raw  <- readxl::read_excel(lm_path)

name_col <- intersect(c("Landmark Name", "Landmark"), names(lm_raw))[1]
if (is.na(name_col)) {
  stop("Landmark file must include either 'Landmark Name' or 'Landmark' plus 'Latitude' and 'Longitude'.")
}

landmarks <- lm_raw %>%
  dplyr::rename(
    name = !!name_col,
    lat  = Latitude,
    lon  = Longitude
  ) %>%
  # Apply label tweaks for annotations
  dplyr::mutate(
    name = name %>%
      stringr::str_replace_all("\\bEl Cerrito\\b", "EC") %>%     # e.g., "El Cerrito Library" -> "EC Library"
      stringr::str_replace("^Harding Elementary$", "Harding") %>% # exact match
      stringr::str_replace("^Madera Elementary School$", "Madera")
  )

# Default label offsets if they don't exist
if (!"hjust" %in% names(landmarks)) landmarks$hjust <- -0.1
if (!"vjust" %in% names(landmarks)) landmarks$vjust <- -0.5

# Filter incidents for the period and keep only rows with coords; drop BAD_LATLON_FLAG if present
heat_data <- df_filtered_clean_updated %>%
  dplyr::filter(!is.na(received_date)) %>%
  dplyr::filter(dplyr::between(lubridate::year(received_date), 2017, 2025)) %>%
  dplyr::filter(!is.na(latitude), !is.na(longitude))

if ("BAD_LATLON_FLAG" %in% names(heat_data)) {
  heat_data <- dplyr::filter(heat_data, !BAD_LATLON_FLAG)
}

# Heat map
ggplot2::ggplot() +
  ggplot2::stat_density_2d(
    data = heat_data,
    ggplot2::aes(x = longitude, y = latitude, fill = after_stat(level)),
    geom = "polygon", contour = TRUE, alpha = 0.8, bins = 40
  ) +
  ggplot2::scale_fill_gradient(low = "lightgreen", high = "red", name = "Incident Density") +
  ggplot2::geom_point(
    data = landmarks,
    ggplot2::aes(x = lon, y = lat),
    color = "black", size = 3, shape = 21, fill = "white"
  ) +
  ggplot2::geom_text(
    data = landmarks,
    ggplot2::aes(x = lon, y = lat, label = name, hjust = hjust, vjust = vjust),
    size = 2.5, fontface = "bold"
  ) +
  ggplot2::labs(
    title = "El Cerrito Police Incidents – Heat Map",
    subtitle = "Filtered incident categories with coordinates (2017–2025)"
  ) +
  ggplot2::coord_fixed() +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.title = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = ggplot2::element_text(hjust = 0.5)
  )



```


```{r save-heatmap-jpeg, eval=TRUE}
# Save heatmap directly to JPEG in project folder for Nextdoor upload with white background
jpeg("D:/Documents/Employment/2025 job search/Project 2025/el-cerrito-police-report/EC-police_incidents.jpg",
     width = 8, height = 6, units = "in", res = 300, bg = "white")

ggplot2::ggplot() +
  ggplot2::stat_density_2d(
    data = heat_data,
    ggplot2::aes(x = longitude, y = latitude, fill = after_stat(level)),
    geom = "polygon", contour = TRUE, alpha = 0.8, bins = 40
  ) +
  ggplot2::scale_fill_gradient(low = "lightgreen", high = "red", name = "Incident Density") +
  ggplot2::geom_point(
    data = landmarks,
    ggplot2::aes(x = lon, y = lat),
    color = "black", size = 3, shape = 21, fill = "white"
  ) +
  ggplot2::geom_text(
    data = landmarks,
    ggplot2::aes(x = lon, y = lat, label = name, hjust = hjust, vjust = vjust),
    size = 2.5, fontface = "bold"
  ) +
  ggplot2::labs(
    title = "El Cerrito Police Incidents – Heat Map",
    subtitle = "Filtered incident categories with coordinates (2017–2025)"
  ) +
  ggplot2::coord_fixed() +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.background = ggplot2::element_rect(fill = "white", colour = "white"), # added white background
    plot.background = ggplot2::element_rect(fill = "white", colour = "white"),  # added white background
    panel.grid = ggplot2::element_blank(),
    axis.title = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = ggplot2::element_text(hjust = 0.5)
  )

dev.off()






```




\newpage


```{r table1a_total_incidents}
# Purpose: Flextable of incident counts within 500 ft of each landmark (rows = landmarks), sorted desc.

# Guards
stopifnot(all(c("longitude","latitude") %in% names(df_filtered_clean_updated)))
stopifnot(exists("landmarks"))
stopifnot(all(c("name","lat","lon") %in% names(landmarks)))

# Prepare incident coordinates once
inc_coords <- df_filtered_clean_updated |>
  dplyr::filter(!is.na(longitude), !is.na(latitude)) |>
  dplyr::select(longitude, latitude) |>
  as.matrix()

radius_m <- 500 * 0.3048

# Fast counter for one landmark
count_near <- function(lat, lon) {
  sum(geosphere::distHaversine(inc_coords, c(lon, lat)) <= radius_m)
}

# Build the 2-column table
landmark_counts <- landmarks |>
  dplyr::mutate(
    `Incident Count (within 500 ft)` = vapply(
      seq_len(n()),
      function(i) count_near(lat[i], lon[i]),
      integer(1)
    )
  ) |>
  dplyr::transmute(
    Landmark = name,
    `Incident Count (within 500 ft)`
  ) |>
  dplyr::arrange(dplyr::desc(`Incident Count (within 500 ft)`))

# Flextable
# Purpose: Use table_fire_style safely for the landmark 500 ft counts  # added
table_fire_style(
  landmark_counts,
  label_map = c(
    Landmark = "Landmark",
    `Incident Count (within 500 ft)` = "Incidents within 500 ft"
  ),
  caption = "Incident Counts within 500 ft of Each Landmark"
)


```




\newpage
```{r table1b_library_incident_types}
# Purpose: Compare incident counts by type for New vs Current Library sites (within 500 ft),
#          and render with the standard table style used in the fire report.           # added

# --- Helpers (same as before) ---
ft_to_m <- function(ft) { ft * 0.3048 }                                              # added

incidents_within_radius <- function(data, center_lat, center_lon, radius_ft = 500) { # added
  valid <- data %>% dplyr::filter(!is.na(latitude), !is.na(longitude))
  if (nrow(valid) == 0) { return(valid[0, ]) }
  d_m <- geosphere::distHaversine(
    matrix(c(valid$longitude, valid$latitude), ncol = 2),
    c(center_lon, center_lat)
  )
  valid %>%
    dplyr::mutate(distance_m = d_m) %>%
    dplyr::filter(distance_m <= ft_to_m(radius_ft))
}

# Search helper: try both name_raw (if present) and name (label-edited)                 # added
find_landmark <- function(patterns, landmarks_df) {                                    # added
  cols_to_search <- intersect(c("name_raw", "name"), names(landmarks_df))
  if (length(cols_to_search) == 0) { stop("Landmarks must have a 'name' column.") }

  for (pat in patterns) {
    for (col in cols_to_search) {
      idx <- grep(pat, landmarks_df[[col]], ignore.case = TRUE)
      if (length(idx) >= 1) {
        if (length(idx) > 1) {
          message("Multiple landmarks matched '", pat, "' via column '", col, "'. Using: ", landmarks_df$name[idx[1]])
        }
        return(landmarks_df[idx[1], c("name", "lat", "lon"), drop = FALSE])
      }
    }
  }

  stop(
    "Landmark not found for patterns: ",
    paste(patterns, collapse = " | "),
    "\nAvailable names:\n- ",
    paste(landmarks_df[[cols_to_search[1]]], collapse = "\n- ")
  )
}

# --- Get landmarks (patterns cover both EC and El Cerrito forms) ---
newlib_patterns     <- c("new.*library", "proposed.*library", "plaza.*library", "ec.*plaza.*library")
currentlib_patterns <- c("current.*library", "ec.*library", "el\\s*cerrito\\s*library", "stockton")

newlib_row     <- find_landmark(newlib_patterns, landmarks)
currentlib_row <- find_landmark(currentlib_patterns, landmarks)

# --- Incidents for each site (500 ft) ---
newlib_df     <- incidents_within_radius(df_filtered_clean_updated, newlib_row$lat, newlib_row$lon, 500)
currentlib_df <- incidents_within_radius(df_filtered_clean_updated, currentlib_row$lat, currentlib_row$lon, 500)

# --- Summary counts (wide) ---
tbl_library_compare <- dplyr::full_join(
  newlib_df %>% dplyr::count(call_for_service, name = "New Library Site"),
  currentlib_df %>% dplyr::count(call_for_service, name = "Current Library Site"),
  by = "call_for_service"
) %>%
  tidyr::replace_na(list(`New Library Site` = 0, `Current Library Site` = 0)) %>%
  dplyr::mutate(
    `New Library Site`     = as.integer(`New Library Site`),                      # added
    `Current Library Site` = as.integer(`Current Library Site`)                   # added
  ) %>%
  dplyr::arrange(dplyr::desc(`New Library Site`))

# --- Styled flextable using your standard function ---
table_fire_style(                                                                     # added
  tbl_library_compare,
  label_map = c(
    call_for_service      = "Incident Type",
    `New Library Site`     = "New Library Site",
    `Current Library Site` = "Current Library Site"
  ),
  caption = "Incident Counts by Type — New vs Current Library Sites (within 500 ft)", # added
  footer  = "Counts within 500 ft of each library site; dataset reflects filtered incident categories." # added
)

```

\newpage

**Heatmap of Car Realted Incidents**  

There is considerable community interest in car related incidents, so I created a heatmap of car related incidents and a table of counts by landmark.  
Please see the appendix for a list of car incident types.

```{r chart2_heatmap_car_incidents}
# Purpose: Heat map of "car incidents" (full dataset filtered to specific codes).
# Uses Excel landmarks with the same label tweaks (EC, Harding, Madera).

# --- Define the car-related call_for_service codes (lowercased to match your data cleaning) ---
car_codes <- c(
  "459a - auto burglary",
  "10851 - motor vehicle theft",
  "10851r - recovered stolen vehicle",
  "215 - car jacking",
  "23110 - throwing objects at a vehicle",
  "10852 - vehicle parts theft"   # (typo matches your dataset)
)

# --- Subset the full dataset to car incidents and valid coordinates ---
car_df <- df_filtered_clean_updated %>%
  dplyr::mutate(call_for_service = stringr::str_to_lower(stringr::str_trim(call_for_service))) %>%
  dplyr::filter(call_for_service %in% car_codes) %>%
  dplyr::filter(!is.na(latitude), !is.na(longitude))

if ("BAD_LATLON_FLAG" %in% names(car_df)) {
  car_df <- dplyr::filter(car_df, !BAD_LATLON_FLAG)
}

# --- Landmarks from Excel (authoritative) + label tweaks ---
lm_path <- "D:/Documents/Employment/2025 job search/Project 2025/el-cerrito-police-report/data/combined/landmark lat and long.xlsx"
lm_raw  <- readxl::read_excel(lm_path)

name_col <- intersect(c("Landmark Name", "Landmark"), names(lm_raw))[1]
if (is.na(name_col)) stop("Landmark file must include either 'Landmark Name' or 'Landmark' plus 'Latitude' and 'Longitude'.")

landmarks <- lm_raw %>%
  dplyr::rename(name = !!name_col, lat = Latitude, lon = Longitude) %>%
  dplyr::mutate(
    # label tweaks you requested
    name = name |>
      stringr::str_replace_all("\\bEl Cerrito\\b", "EC") |>
      stringr::str_replace("^Harding Elementary$", "Harding") |>
      stringr::str_replace("^Madera Elementary School$", "Madera")
  )

# Default label offsets if not present
if (!"hjust" %in% names(landmarks)) landmarks$hjust <- -0.1
if (!"vjust" %in% names(landmarks)) landmarks$vjust <- -0.5

# --- Heat map ---
ggplot2::ggplot() +
  ggplot2::stat_density_2d(
    data  = car_df,
    ggplot2::aes(x = longitude, y = latitude, fill = after_stat(level)),
    geom  = "polygon",
    contour = TRUE,
    alpha = 0.8,
    bins  = 40
  ) +
  ggplot2::scale_fill_gradient(low = "lightgreen", high = "red", name = "Incident Density") +
  ggplot2::geom_point(
    data = landmarks,
    ggplot2::aes(x = lon, y = lat),
    color = "black", size = 3, shape = 21, fill = "white"
  ) +
  ggplot2::geom_text(
    data = landmarks,
    ggplot2::aes(x = lon, y = lat, label = name, hjust = hjust, vjust = vjust),
    size = 2.5, fontface = "bold"
  ) +
  ggplot2::labs(
    title = "Car Incidents — Heat Map",
    subtitle = "459a, 10851, 10851r, 215, 23110, 10852 (full dataset)"
  ) +
  ggplot2::coord_fixed() +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid   = ggplot2::element_blank(),
    axis.title   = ggplot2::element_blank(),
    axis.text    = ggplot2::element_blank(),
    axis.ticks   = ggplot2::element_blank(),
    plot.title   = ggplot2::element_text(hjust = 0.5, face = "bold"),
    plot.subtitle= ggplot2::element_text(hjust = 0.5)
  )

```

\newpage

```{r table2_car_incidents}
# Purpose: Flextable — counts of "car incidents" within 500 ft of each landmark, sorted descending.
# Style: Uses table_fire_style() to match the fire report formatting.
# Notes:
# - Car incidents are limited to the specified call_for_service codes (lowercased to match cleaned data)
# - Reuses `car_df` and `landmarks` if they already exist

# --- Ensure car_df exists (only build if missing) ---
if (!exists("car_df")) {
  car_codes <- c(
    "459a - auto burglary",
    "10851 - motor vehicle theft",
    "10851r - recovered stolen vehicle",
    "215 - car jacking",
    "23110 - throwing objects at a vehicle",
    "10852 - vehcile parts theft"  # keep the source-data spelling for matching
  )

  car_df <- df_filtered_clean_updated %>%
    dplyr::mutate(call_for_service = stringr::str_to_lower(stringr::str_trim(call_for_service))) %>%
    dplyr::filter(call_for_service %in% car_codes) %>%
    dplyr::filter(!is.na(latitude), !is.na(longitude))

  if ("BAD_LATLON_FLAG" %in% names(car_df)) {
    car_df <- dplyr::filter(car_df, !BAD_LATLON_FLAG)
  }
}

# --- Ensure landmarks exists (only load if missing) ---
if (!exists("landmarks")) {
  lm_path <- file.path(
    "D:/Documents/Employment/2025 job search/Project 2025/el-cerrito-police-report/data/combined",
    "landmark lat and long.xlsx"
  )
  lm_raw  <- readxl::read_excel(lm_path)

  name_col <- intersect(c("Landmark Name", "Landmark"), names(lm_raw))[1]
  if (is.na(name_col)) {
    stop("Landmark file must include either 'Landmark Name' or 'Landmark' plus 'Latitude' and 'Longitude'.")
  }

  landmarks <- lm_raw %>%
    dplyr::rename(name = !!name_col, lat = Latitude, lon = Longitude) %>%
    dplyr::mutate(
      name = name |>
        stringr::str_replace_all("\\bEl Cerrito\\b", "EC") |>
        stringr::str_replace("^Harding Elementary$", "Harding") |>
        stringr::str_replace("^Madera Elementary School$", "Madera")
    )

  if (!"hjust" %in% names(landmarks)) landmarks$hjust <- -0.1
  if (!"vjust" %in% names(landmarks)) landmarks$vjust <- -0.5
}

# --- Guards ---
stopifnot(all(c("longitude", "latitude") %in% names(car_df)))
stopifnot(all(c("name", "lat", "lon") %in% names(landmarks)))

# --- Distance helper ---
radius_m <- 500 * 0.3048  # 500 ft in meters

inc_coords <- as.matrix(car_df[, c("longitude", "latitude")])

count_near <- function(lat, lon) {
  if (!nrow(car_df)) return(0L)
  sum(geosphere::distHaversine(inc_coords, c(lon, lat)) <= radius_m)
}

# --- Build counts table ---
car_counts <- landmarks %>%
  dplyr::mutate(
    `Incidents within 500 ft` = vapply(seq_len(n()), function(i) count_near(lat[i], lon[i]), integer(1))
  ) %>%
  dplyr::transmute(Landmark = name, `Incidents within 500 ft`) %>%
  dplyr::arrange(dplyr::desc(`Incidents within 500 ft`))

# --- Styled flextable (no commas on counts) ---
table_fire_style(
  car_counts,
  label_map = c(
    Landmark = "Landmark",
    `Incidents within 500 ft` = "Incidents within 500 ft"
  ),
  caption = "Car Incidents within 500 ft of Each Landmark (descending)",
  footer  = "Includes: 459a, 10851, 10851r, 215, 23110, 10852. Distances computed from landmark coordinates."
)

```

\newpage

**Incidents by Year**  

Chart 3 clearly shows the reduced incident counts during the Covid years of 2020-2022. During those years the police department was short-staffed, so it is not clear if crime was down or the ability of the department was diminished or both. I leave it to others to investigate that point. 

```{r chart3_line__total_incidents}
# Purpose: Line chart of total incidents by year, with note for 2025 half-year.

yearly_totals <- df_filtered_clean_updated |>
  dplyr::filter(!is.na(received_date)) |>
  dplyr::mutate(Year = lubridate::year(received_date)) |>
  dplyr::count(Year, name = "Incidents")

ggplot2::ggplot(yearly_totals, ggplot2::aes(x = Year, y = Incidents)) +
  ggplot2::geom_line(linewidth = 1.2, color = "blue") +
  ggplot2::geom_point(size = 2, color = "blue") +
  ggplot2::geom_text(
    data = subset(yearly_totals, Year == 2025),
    aes(label = "Only six months of data"),
    hjust = 1.2, vjust = -0.5, size = 3   # hjust > 1 moves text to the left
  ) +
  ggplot2::scale_x_continuous(breaks = yearly_totals$Year) +
  ggplot2::labs(
    title = "Total Incidents by Year",
    x = "Year", y = "Incidents"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(hjust = 0.5, face = "bold")
  )
```


\newpage


```{r table3_counts_by_year}

# Purpose: Total incidents per year with standard styling, no commas on counts, and a footer if latest year is partial.
# Notes:
# - Uses table_fire_style() for consistent look (fire report style).
# - YEAR and Incidents shown without thousands separators per your preference.

# --- Recompute yearly counts (simple, self-contained) ---
yearly_counts_tbl <- df_filtered_clean_updated %>%
  dplyr::filter(!is.na(received_date)) %>%
  dplyr::mutate(YEAR = lubridate::year(received_date)) %>%
  dplyr::count(YEAR, name = "Incidents") %>%
  dplyr::arrange(YEAR)

# --- Detect partial latest year (<= June) ---
latest_year <- max(lubridate::year(df_filtered_clean_updated$received_date), na.rm = TRUE)
max_month_latest <- df_filtered_clean_updated %>%
  dplyr::filter(!is.na(received_date), lubridate::year(received_date) == latest_year) %>%
  dplyr::summarise(m = max(lubridate::month(received_date), na.rm = TRUE), .groups = "drop") %>%
  dplyr::pull(m)
is_partial <- !is.na(max_month_latest) && max_month_latest <= 6

# --- Base styled flextable ---
ft <- table_fire_style(
  yearly_counts_tbl,
  label_map = c(YEAR = "Year", Incidents = "Incidents"),
  caption   = "Total Police Incidents by Year"
)

# --- Formatting tweaks: remove commas from counts; YEAR already handled in the helper ---
ft <- flextable::colformat_num(ft, j = "Incidents", digits = 0, big.mark = "")  # no commas

# --- Footer note if partial latest year ---
if (is_partial) {
  ft <- flextable::add_footer_lines(ft, paste0(latest_year, " is Jan–Jun only."))
}

ft



```


\newpage

**Incidents by Month**   

Chart 4 is incident by month. Since I only had data for the first six months of 2025, I did not use the 2025 data for this chart. The chart shows that incident counts are fairly similar throughout the year.  


```{r chart4_monthly_counts}

# Purpose: Bar chart of total incidents by month (2019–2024), with labeled average line.
# Ensures all 12 months appear (uses tidyr::complete).

df_19_24 <- df_filtered_clean_updated |>
  dplyr::filter(!is.na(received_date)) |>
  dplyr::filter(lubridate::year(received_date) >= 2019,
                lubridate::year(received_date) <= 2024) |>
  dplyr::mutate(
    MonthNum = lubridate::month(received_date),
    Month    = factor(month.name[MonthNum], levels = month.name)
  ) |>
  dplyr::count(Month, name = "Incidents") |>
  tidyr::complete(Month = factor(month.name, levels = month.name),
                  fill = list(Incidents = 0)) |>
  dplyr::arrange(Month)

monthly_totals <- df_19_24  # keep for the table chunk

avg_month <- mean(monthly_totals$Incidents, na.rm = TRUE)

ggplot2::ggplot(monthly_totals, ggplot2::aes(x = Month, y = Incidents)) +
  ggplot2::geom_col(fill = "#cfe8ff", color = "black", width = 0.8) +
  ggplot2::geom_hline(yintercept = avg_month, linetype = "dashed", linewidth = 0.9) +
  ggplot2::annotate(
    "text",
    x = length(month.name) - 0.2,
    y = avg_month,
    label = paste0("Average (2019–2024): ", round(avg_month, 0)),
    vjust = -0.6, size = 3
  ) +
  ggplot2::labs(
    title = "Total Incidents by Month (2019–2024)",
    x = NULL, y = "Incidents"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid.major.x = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(hjust = 0.5, face = "bold")
  )

```

\newpage


```{r table4_monthly_counts}

# Purpose: Monthly incident totals (2019–2024) with an average row, styled like the fire report
# Notes:
# - Reuses `monthly_totals` from the chart if present; otherwise rebuilds
# - Removes thousands separators on counts
# - Adds a footer noting 2025 exclusion

# --- Rebuild if needed (keep order Jan–Dec) ----------------------------------------- # added
if (!exists("monthly_totals")) {
  monthly_totals <- df_filtered_clean_updated |>
    dplyr::filter(!is.na(received_date)) |>
    dplyr::filter(lubridate::year(received_date) >= 2019,
                  lubridate::year(received_date) <= 2024) |>
    dplyr::mutate(
      MonthNum = lubridate::month(received_date),
      Month    = factor(month.name[MonthNum], levels = month.name)                  # added
    ) |>
    dplyr::count(Month, name = "Incidents") |>
    tidyr::complete(Month = factor(month.name, levels = month.name),
                    fill = list(Incidents = 0)) |>
    dplyr::arrange(Month)
}

# --- Average row -------------------------------------------------------------------- # added
avg_row <- tibble::tibble(
  Month = "Average (2019–2024)",
  Incidents = round(mean(monthly_totals$Incidents, na.rm = TRUE), 0)
)

monthly_totals_with_avg <- dplyr::bind_rows(monthly_totals, avg_row)

# --- Styled flextable --------------------------------------------------------------- # added
ft <- table_fire_style(
  monthly_totals_with_avg,
  label_map = c(Month = "Month", Incidents = "Incidents"),
  caption   = "Total Incidents by Month (2019–2024)",
  footer    = "Note: 2025 excluded (only half year available)."
)

# Remove commas from counts (keep Year/Month text as-is)                               # added
ft <- flextable::colformat_num(ft, j = "Incidents", digits = 0, big.mark = "")

ft



```

\newpage

## Appendix  

### Technical Note  

I originally received the El Cerrito Police Department data in a PDF format, so I used Excel/Get Data (aka Power Query) to read the data into Excel. There was further clean up with Power Query.  
Then I read the data into R. I use RStudio and the tidyverse.  
I then made decisions on what to exclude. I excluded data that did not have a full address. In particular addresses such as “San Pablo Avenue” were excluded. I also limited the types of incidents to the ones listed below.  

### Included Incident Types  
These are the included incident types. The notation comes straight from the El Cerrito Police Department data.  
Incident Types  
unwant - unwanted person  
415 - disturbance  
488 - petty theft  
parker - parking violation  
1154 - suspicious vehicle  
459a - auto burglary  
10851 - motor vehicle theft  
594 - vandalism  
1053 - person down  
243a - assault / battery  
487 - grand theft  
211 - robbery  
417 - brandishing  
1179 - accident w/ medical routed  
422 - criminal threats  
314 - indecent exposure  
hs - narcotics use/possession  
10851r - recovered stolen vehicle  
215 - car jacking  
23110 - throwing objects at a vehicle  
459c - commercial burglary  
459r - residential burglary  
fight - physical fight  
sfrmc - shots fired  
211a - business hold up/robbery alarm  
10852 - vehcile parts theft  
dopers - narcotic sales  
1027 - wanted person / warrant  
148.1 - bomb threat  
stalk - stalking  
496 - stolen property  
245 - assault with a deadly weapon  
1071 - shooting (actual victim)  
1051 - drunk / intoxicated  
207 - kidnapping  
mgun - person with a gun  
loiter - loitering  
459 - burglary (misc)  
1070 - prowler  
647b - prostitution  
243.4 - sexual battery / assault  
370 - public nuisance  
22810 - assault with pepper spray / tear gas  
246 - shooting into a dwelling  
247b - shooting into an unoccupied vehicle  
220 - assault with intent to rape or rob  
1075 - shots fired (shotspotter)  

### Automobile Related Incident Types  
459a - auto burglary  
10851 - motor vehicle theft  
10851r - recovered stolen vehicle  
215 - car jacking 
23110 - throwing objects at a vehicle  
10852 - vehicle parts theft  

